Tree height and hydraulic traits shape growth responses across droughts in a temperate broadleaf forest

### 2020-07-24 manuscript/response_to_reviews.Rmd

**Thank you.**

349 fascinating; it seems plausible that current climate changed induced large-individual mortality involves trees growing to conduit diameters, and thus heights, permitted under previously more even and moister climates, and now finding themselves in more extreme climates that make their heights and conduits excessively wide given current conditions.

**We have added the following text:**

**This may be because moister habitats would tend to support species and individuals with more mesophytic traits [@bartlett_drought_2016; @mencuccini_ecological_2003; @medeiros_extensive_2019], potentially growing to greater heights [e.g., @detto_hydrological_2013], and these are then more vulnerable when drought hits.**

@@

I enjoyed the paper very much. It’s clearly and succinctly written, and brings together a range of complementary datasets to address some very topical questions in forest ecology. I could definitely see the paper being of interest to a broad readership. However, I do have some important questions and suggestions about the analysis which I think need more careful thought. Below I’ve listed my main concerns, following which there are some other comments that the authors may want to take into consideration in revising their paper.

Main comments

L139-140: makes sense to not core live trees in the plot (surprised you got permission to do this in 2011!). But doesn’t this introduce a bias in your analysis? Trees that died may have had very different responses to drought that those which survived. Do you get similar results if you analyse trees that lived and died separately?

**We agree that this could be a concern, but do not believe it is applicable here, for two reasons. First, this would be more likely to introduce a bias if the trees cored dead were killed by drought. We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived. However, these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years. Second, quoting our previous publication using these cores [@helcoski_growing_2019]: "Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.com/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/results/live_vs_dead).".**

@@

**We do not agree that the $Rt$ metric "assumes that tree growth (BAI in this case) fluctuates around a long term mean that is stable through time." If this was true, $Rt$ would be calculated by dividing growth (BAI) in the drought year by the long-term mean growth increment. Instead, $Rt$ is calculated by dividing growth in the focal year by the mean growth over a narrow window of preceding years (5 years in this study). Even though there may be a long-term directional trend in BAI, if you were to calculate the trend in BAI over each 5-year increment, the trend could go in any direction. To account for such trends, we tried an intervention time series analysis (ARIMA). Details on the comparison between $Rt$ and $Rt_{ARIMA}$ are presented in the responses to reviewer 4. In short, the ARIMA method yielded broadly similar results, but tended to produce less reasonable estimates when the difference between the two metrics was large, so we decided to retain $Rt$ as the focal metric. We now include the following content on ARIMA in the manuscript:**

**"Because the $Rt$ metric could be biased by directional pre-drought growth trends, we also tried an intervention time series analysis [ARIMA, @R-forecast] that predicted mean drought-year growth based on trends over the preceding ten years and used this value in place of the five-year mean in calculations of resistance ($Rt_{ARIMA}$= observed $BAI$/ predicted $BAI$). The two metrics were strongly correlated (Fig. **S5**). Visual review of the individual tree-ring sequences with the largest discrepancies between these metrics revealed that $Rt$ was less prone to unreasonable estimates than $Rt_{ARIMA}$, so we selected $Rt$ as our focal metric, presenting parallel results for $Rt_{ARIMA}$ in the Supplementary Info."**

@@

**"Because the resistance and resilience of growth to drought is linked to their probability of surviving drought [@desoto_low_2020; @liu_hydraulic_2019], understanding growth responses can also help elucidate which trees are most vulnerable to drought-induced mortality."**

L80-101 & L112-115: This is a nice summary of how traits can help us understand species responses to drought. The only thing that is a little strange is that you start by discussing commonly measured traits that have been used to explore susceptibility to drought (e.g., WD and LMA). Then you say that the low predictive power of these traits is because they are poor surrogates of what we should be measuring – i.e. hydraulic traits. But then you end by saying that these are too hard/expensive to measure, so you’re going to look at other surrogates.

**We have revised this section, and believe that it now makes our points more clearly: (1) the traditionally measured traits are poor surrogates, (2) hydraulic traits are useful but time-consuming, (3) $PLA_{dry}$ and $\pi_{tlp}$ are promising alternatives that are both relatively easy to measure and have direct physiological linkages to tree growth and mortality responses to drought. The text now reads as follows:**

**"Debates have also arisen regarding the traits influencing tree growth responses to drought. Studies within temperate broadleaf forests have observed ring-porous species showing higher drought tolerance than diffuse-porous species [@friedrichs_species-specific_2009; @elliott_forest_2015; @kannenberg_linking_2019], but this distinction would not hold in the global context [@wheeler_variations_2007; @olson_xylem_2020] and does not resolve differences among the many species within each category. Commonly-measured traits including wood density and leaf mass per area ($LMA$) have been linked to drought responses within some temperate deciduous forests [@abrams_adaptations_1990; @guerfel_impacts_2009; @hoffmann_hydraulic_2011; @martinbenito_convergence_2015] and across forests worldwide [@greenwood_tree_2017]. However, in other cases these traits could not explain drought tolerance [e.g., in a tropical rainforest; @marechaux_leaf_2019], or the direction of response was not always consistent. For instance, higher wood density has been associated with greater drought resistance at a global scale [@greenwood_tree_2017], but correlated negatively with tree performance during drought in a broadleaf deciduous forest in the southeastern United States [@hoffmann_hydraulic_2011]. Thus, the perceived influence of these traits on drought resistance may actually reflect indirect correlations with other traits that more directly drive drought responses [@hoffmann_hydraulic_2011]."**

@@

L230: Somewhere here it would be good to remind the reader what these explanatory variables are.

**We have revised this sentence to read, "The general statistical model for hypothesis testing was a mixed effects model, implemented  in the lme4 package in R [@R-lme4], using tree nested within species as a random effect, and independent variables as specified below." We later specify the list of variables.**

L234-237: Not very clear how these models were constructed. Why did you fit each predictor independently to test its importance as opposed to standardizing them and having them together in a multiple regression and then compare their coefficients? And why was height retained in all models? Testing one predictor at a time might be sensible if you have lots of collinearity between predictors. But it could also mean you miss important conditional effects (e.g., the effect of variable x only emerges when you first account for the effect of variable y).

**We have revised our method, and now present a single main model with all variables at once. However, we did conduct preliminary independent tests of each species trait to avoid over-fitting in the model. The most relevant text now reads:**

**"To avoid over-fitting models with five species traits (Table 3) across only 12 species, we did not include all traits as fixed effects in a single linear mixed model, but rather conducted individual tests of each species trait to determine the relative importance and appropriateness for inclusion in the main model. These tests followed the model structure specified above, then added $ln[H]$ and $ln[TWI]$ to create a base model against which we tested traits. Trait variables were considered appropriate for inclusion in the main model if they had a consistent direction of response across all droughts and if their addition to the base model improved fit (at $\Delta$AICc $\ge$ 1.0) in at least one drought year (Table S4). We note that we did not use the $\Delta$AICc $\ge$ 1.0 criterion as a test of significance, but rather of whether the variable had enough influence to be considered as a *candidate* variable in full models."**

@@

L255-258: good to have this summary, but it’s sort of expected that you have a decline in growth as this is how you defined drought (a little circular). So given that you’ve already mentioned how you defined drought in the methods maybe you don’t need to repeat this here?

**We note that the intent was never to test for a decline in growth, but to examine how it varied across trees; hence, we do not see it as circular. However, we have dropped the criteria of decline in growth as part of how we defined drought. This does not change the drought years that qualify for inclusion.**

@@

L67: I suggest toning down the sentence to something like ‘Wider xylem conduits are associated to higher vulnerability to embolism during drought’, as the mechanism linking wider conduits with embolism vulnerability is far from clear and probably indirect.

**We have revised this to read, "Wider xylem conduits plausibly make large trees more vulnerable to embolism during drought [@olson_plant_2018]..."**

L81: you seem to use drought resistance and drought tolerance interchangeably, which may be confusing. I think it is important to clarify these definitions early on in the paper.

**We now define drought resistance in the first paragraph: "Understanding forest responses to drought requires elucidation of how tree size, microenvironment, and species’ traits jointly influence individual-level drought resistance, defined here as a tree's ability to maintain growth during drought,...". We then use this term when we wish to speak specifically about resistance. We use "tolerance" as a more general term, consistent with what seems to be a fairly nebulous definition in the literature.**

@@

L100-101: that is not true really. See for example Zhu et al. (2018) in Tree Phys, Farrell et al. (2017) in PCE…

**We have added citations to both these papers and revised the sentence so that it is true: "The abilities of both $PLA_{dry}$ and $\pi_{tlp}$ to explain tree drought resistance remains untested."**

L137-140: mixing cores from dying and surviving individuals could be problematic. Did you find any association between survival and previous drought resistance? Was mortality different across species?

**We agree that this could be a concern, but do not believe it is applicable here, for two reasons. First, this would be more likely to introduce a bias if the trees cored dead were killed by drought. We are aware of studies showing differences in climate sensitivity of trees killed by drought versus those that survived. However,these trees were not killed by drought; there was no significant drought in the years in which dead trees were cored or in the 3 previous years. Second, quoting our previous publication using these cores [@helcoski_growing_2019]: "Chronologies of trees cored live and dead were pooled following analyses showing similar climate sensitivity at least up to 2009 (that is excluding 7–8 yr before death) for all four species with $\ge$ 10 cores in each category (LITU, QURU, QUVE, FRAM; comparison plots available at https://github.c om/SCBI-ForestGEO/climate_sensitivity_cores/tree/master/ results/live_vs_dead).".**

### 2020-07-24 manuscript/climate_sensitivity_SI.Rmd
checked SI

\listoftables
\listoffigures
\newpage

@@

\newpage
## Further Package Citations
While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes. As in all of science, this study is a representation of the work done by both the authors of this paper as well as countless others. While acknowledging everyone is impossible, we want to at least give thanks to those who made this work possible.

>>>>

\listoftables
\listoffigures

**Appendix S1. Further Package Citations**

@@

\newpage
## Appendix S1. Further Package Citations
While there were several R-packages we used for a specific purpose in our methods, numerous packages were immensely helpful for this research behind the scenes. As in all of science, this study is a representation of the work done by both the authors of this paper as well as countless others. While acknowledging everyone is impossible, we want to at least give thanks to those who made this work possible.

### 2020-07-24 manuscript/climate_sensitivity_draft.Rmd
checked SI

All analysis beyond basic data collection was performed using R version 3.6.2 [@R-base]. 
Other R-packages used in analyses are listed in the Supplementary Information (*Appendix S1*).
All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.

@@

Table S1: Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.

Table S2: Species-specific bark thickness regression equations.

Table S3: Species-specific height regression equations.

@@

Figure S2: Map of ForestGEO plot showing topographic wetness index and location of cored trees

Figure S3: Distribution of reconstructed tree heights across drought years.

@@

>>>>

All analysis beyond basic data collection was performed using R version 3.6.2 [@R-base]. 
Other R-packages used in analyses are listed in the Supplementary Information (Appendix S1).
All data, code, and results are available through the SCBI-ForestGEO organization on GitHub (https://github.com/SCBI-ForestGEO: SCBI-ForestGEO-Data and McGregor_climate-sensitivity-variation repositories), with static versions corresponding to data and analyses presented here archived in Zenodo (DOIs: 10.5281/zenodo.3604993 and *[TBD]*, respectively.

@@

Table S1. Monthly Palmer Drought Severity Index (PDSI), and its rank among all years between 1950 and 2009 (driest=1), for focal droughts.

Table S2. Species-specific bark thickness regression equations.

Table S3. Species-specific height regression equations.

@@

Figure S2. Map of ForestGEO plot showing topographic wetness index and location of cored trees

Figure S3. Distribution of reconstructed tree heights across drought years.

@@

Appendix S1. Further Package Citations
